Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-07-19 review_article.Rmd

- Future steps should consider applications not currently utilised for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualising those spatial patterns, or the uncertainty in those patterns
- a more applied-focussed paper (eg lots of cool images), rather than theoretical (eg lots of formula).
-->

@@

Utilising maps to present statistics has been widely used for centuries.
Connecting the data to the geographical representation of areas that are already familiar.
It is not enough to be recognisable, if the value of the statistic cannot be seen. Cancer outcomes directly relate to the people living within a geographical area. In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer outcomes. This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes.

@@

<!-- psychology -->
Utilising the familiar state shapes can make maps intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data.
The value of the familiarity of the geography is worth consideration when presenting results of spatial analysis. 
Just as geographers are no longer the only creators of maps, @CPISACA suggests the audience of spatial health data analysis has extended beyond researchers to the public, policymakers and the media.
However, while the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, different population densities and subject to changes over time. 
The different sizes of administrative areas can attract undue attention to the shades of the unpopulated, large physical areas [@EI]. @GOINO believes there are "severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually".
As the display may draw attention from the 'potentially more important results in the more populous communities' [@SE], choropleths can inhibit visual inference when presenting human related statistics.

@@

In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
Almost 100 years of cancer maps have explored the United States and the United Kingdom with increased effectiveness in the presentation of unbiased rates. The increasing development and use of disease maps can be attributed to the availability of location data and associated data points [@GOINO], and geographic information system software [@SE]. Choropleth visualisation suits the data collection methods of census data and cancer mortality rates across regions, and administrative control within regions.

@@

A grey literature review conducted by (ref) identified 33 Cancer Atlases publically available cancer atlases available on the internet and published between 01/01/2010 to 11/11/2015.

All the maps identified in this review were chloropleth maps, in which the representation of the map geometry stayed true to the underlying geography.

<!-- Could be nice to insert a grid of the different maps. Or at least one or two full maps -->

@@

The review did not reveal any novel uncertainty visualisation approaches or visualisations. Maps used standard and well known measures including credible intervals and standard deviation, statistical significance, boxplots and distributions. These maps ranged from static pdfs or infographics to interactive online resources. The interactivity of the more modern maps enabled uncertainty information to be incorporated without cluttering the screen, such as in a tool tip feature.

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, eg population, mortality -->
Waldo Tobler [-@TVSSS] explores many graphical techniques, and suggests there are particular methods for particular purposes. Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
Cartograms may be seen as an extension of map tranformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].

@@

The creation of cartograms was largely in the hands of cartographers. 
@ACTUC discusses early approaches including John Hunter and Johnathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball bearing approach and Tobler's (1973) computer programs.

@@

There is no unique solution to the topologic
problem of preparing a population-by-area
cartogram for a given population and its jurisdictional
boundaries. Although the map areas
of the various regions are fixed by their popu¬
lations, their shapes and relative locations are
not completely determined.

Other characteristics of the population
may also serve as the base for adjustment; for
example, sex, race, parity, socioeconomic level,
or iminune status.

@@

<!--
[-@NAC] observation units stay in more or less their correct location and maintain correct shape, but at least some units are separated by gaps from their neighbors.
Each area in a non-contiguous cartogram is scaled down or up according to value for the specific area.
A scaling factor specific to

@@

<!--
-simple shapes of the circle, square, and triangle are relatively easily recognized.
-sacrifice communication efficiency;
-There is evidence, too, that angular shapes are more easily seeu than ones with smooth outlines. Shapes that contain the qualities of simplicity and reg,ularity are more readily perceived than others that do not possess these qualities.
-Finally, psychologists tell us that shapes having familiarity are more quickly
recognized than nonsense shapes or those that have no meaning.
-The cartographer can oiily efficiently communicate with this kind of map if he has preserved particular elements of the original outline and incorporated them in the new shape.
-
@NISCC 
-->

@@

@TFYCC Value-Area Cartograms. In these cartograms a region,country, or continent is subdivided into small regions, each ofwhich is represented by a rectangle. This rectangle is proportionate in area to the value which it represents in certain statistical distributions. The regions are grouped in approximately the same positions as they are on the map.

@@

With that in mind, the intented user and message to communicate should drive map selection.
first is the nature or inherent properties of the device itself
second is the ease and accuracy with which map users are able to extract information
from it

@@

@ECGC ?? Highlight the value of animating contigous to see changes over time, US can be recognisable but animation aides interpretation

@@

What software did we use, for eg

>>>>

- Future steps should consider applications not currently utilised for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualising those spatial patterns, or the uncertainty in those patterns
- a more applied-focussed paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

@@

Utilising maps to present statistics has been widely used for centuries.
These maps connect data and/or statistics to the geographical representation of areas that are generally familiar to the audience. However, it is not enough to be recognisable, if the meaning of the statistic cannot be seen. This is particularly relevant to cancer statistics, as cancer outcomes directly relate to the people living within a geographical area. In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer statistics. This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes, and health outcomes more broadly.

@@

<!-- psychology -->

Utilising the familiar state boundaries can make a map intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data, i.e. how cancer rate differs across the states.
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.
However, while the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the unpopulated but large areas [@EI]. Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention from the 'potentially more important results in the more populous communities' [@SE].

<!--
Certain limitations to the standard graphic techniques have also become apparent, however.
They are slow, and do not permit instant retrieval of information. They are untrustworthy when interpretations of spatial association are needed; there are both severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually.

These opportunities will be lost if locational coding is not inserted into the data referencing system as such banks are built, however.

of some element, for example population density, or purchasing power, or agricultural fertility; reality, of course, is not like this. the background variable assumed to be uniform (such as population density) can be adjusted so that it behaves as if uniformity were the case
-->

@@

In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Almost 100 years of cancer mapping in the United States and the United Kingdom has increased the effectiveness in the presentation of unbiased rates. The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE]. The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself to choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the the higher prevalence of cancers in older populations. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps used at this time presented levels of cancer via hatchings on a black and white scale.

@@

A grey literature review conducted by (ref) identified 33 publicly available cancer atlases published on the internet between 01/01/2010 to 11/11/2015. Reflecting the high use of chloropleth maps founds in the literature, outlined in the section above, all the public facing maps identified in this review were also chloropleth maps.

@@

The review did not reveal any novel uncertainty visualisation approaches or visualisations. Maps used standard and well known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. These maps ranged from static pdfs or infographics to interactive online resources. The interactivity of the more modern maps enabled uncertainty information to be incorporated without cluttering the screen, such as in a tool tip feature.

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Waldo Tobler [-@TVSSS] explores many graphical techniques, and suggests there are particular methods for particular purposes. Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].

@@

The creation of cartograms was largely in the hands of cartographers. 
@ACTUC discusses early approaches including John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball bearing approach and Tobler's (1973) computer programs.

@@

There is no unique solution to the topology
problem of preparing a population-by-area
cartogram for a given population and its jurisdictional
boundaries. Although the map areas
of the various regions are fixed by their populations, their shapes and relative locations are
not completely determined.

Other characteristics of the population
may also serve as the base for adjustment; for
example, sex, race, parity, socio-economic level,
or immune status.

@@

<!--
[-@NAC] observation units stay in more or less their correct location and maintain correct shape, but at least some units are separated by gaps from their neighbours.
Each area in a non-contiguous cartogram is scaled down or up according to value for the specific area.
A scaling factor specific to

@@

<!--
-simple shapes of the circle, square, and triangle are relatively easily recognized.
-sacrifice communication efficiency;
-There is evidence, too, that angular shapes are more easily seen than ones with smooth outlines. Shapes that contain the qualities of simplicity and regularity are more readily perceived than others that do not possess these qualities.
-Finally, psychologists tell us that shapes having familiarity are more quickly
recognized than nonsense shapes or those that have no meaning.
-The cartographer can efficiently communicate with this kind of map if he has preserved particular elements of the original outline and incorporated them in the new shape.
-
@NISCC 
-->

@@

@TFYCC Value-Area Cartograms. In these cartograms a region,country, or continent is subdivided into small regions, each of which is represented by a rectangle. This rectangle is proportionate in area to the value which it represents in certain statistical distributions. The regions are grouped in approximately the same positions as they are on the map.

@@

With that in mind, the intended user and message to communicate should drive map selection.
first is the nature or inherent properties of the device itself
second is the ease and accuracy with which map users are able to extract information
from it

@@

@ECGC ?? Highlight the value of animating contiguous to see changes over time, US can be recognisable but animation aides interpretation

@@

What software did we use, for e.g.
